---
title: "Working with NOAA Significant Earthquake Data"
author: "Amlan Mishra"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Working with NOAA Significant Earthquake Data}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

library(rEarthquake)
```

# Introduction

The `rEarthquake` package provides tools for working with
NOAA significant earthquake data.

The package provides functions for:

- cleaning earthquake data;
- cleaning earthquake location names;
- creating earthquake timelines;
- adding labels to earthquake timelines;
- mapping earthquake epicenters; and
- creating informative HTML labels for map popups.

# Cleaning earthquake data

The NOAA significant earthquake dataset contains information
about earthquake dates, locations, magnitudes, coordinates,
deaths, and other characteristics.

The `eq_clean_data()` function creates a `DATE` variable,
converts latitude and longitude to numeric values, and cleans
the earthquake location names.

First, we read the NOAA data included with the package.

```{r read-data}
data <- readr::read_delim(
  system.file(
    "extdata",
    "signif.txt",
    package = "rEarthquake"
  ),
  delim = "\t",
  show_col_types = FALSE
)

clean_data <- eq_clean_data(data)

head(
  clean_data[, c(
    "DATE",
    "LOCATION_NAME",
    "LATITUDE",
    "LONGITUDE",
    "EQ_PRIMARY"
  )]
)
```

# Cleaning earthquake locations

The `eq_location_clean()` function removes the country name
and colon from `LOCATION_NAME` and converts the remaining
location name to title case.

For example:

```{r location-clean}
eq_location_clean(
  "JORDAN: BAB-A-DARAA,AL-KARAK"
)
```

# Visualizing earthquakes over time

The `geom_timeline()` function creates a timeline of
earthquakes.

The x-axis represents the earthquake date and the y-axis
can be used to create separate timelines for different
countries.

```{r timeline}
timeline_data <- clean_data |>
  dplyr::filter(
    COUNTRY %in% c("USA", "JAPAN", "CHINA"),
    !is.na(DATE)
  )

ggplot2::ggplot(timeline_data) +
  geom_timeline(
    ggplot2::aes(
      x = DATE,
      y = COUNTRY
    )
  )
```

# Adding earthquake labels

The `geom_timeline_label()` function adds annotations to
selected earthquakes.

The `n_max` argument can be used to label only the largest
earthquakes.

```{r timeline-label}
ggplot2::ggplot(timeline_data) +
  geom_timeline(
    ggplot2::aes(
      x = DATE,
      y = COUNTRY
    )
  ) +
  geom_timeline_label(
    ggplot2::aes(
      x = DATE,
      y = COUNTRY,
      label = LOCATION_NAME
    ),
    n_max = 5
  )
```

# Mapping earthquake epicenters

The `eq_map()` function creates an interactive leaflet map
of earthquake epicenters.

The `annot_col` argument specifies the column that will be
displayed in the popup.

For example, we can select earthquakes in Mexico from the
year 2000 onward.

```{r mexico-data}
mexico_data <- clean_data |>
  dplyr::filter(
    COUNTRY == "MEXICO",
    !is.na(DATE),
    lubridate::year(DATE) >= 2000
  )
```

The following code creates the interactive map.

```{r mexico-map, eval=FALSE}
eq_map(
  mexico_data,
  annot_col = "DATE"
)
```

# Creating informative map labels

The `eq_create_label()` function creates HTML text containing
the earthquake location, magnitude, and total number of deaths.

Missing values are omitted from the corresponding label.

```{r create-label}
mexico_data$popup_text <- eq_create_label(
  mexico_data
)

head(
  mexico_data$popup_text,
  5
)
```

The generated HTML can then be supplied to `eq_map()` as
the popup annotation.

```{r labelled-map, eval=FALSE}
eq_map(
  mexico_data,
  annot_col = "popup_text"
)
```

# Conclusion

The `rEarthquake` package provides a workflow for cleaning
NOAA significant earthquake data and visualizing earthquake
activity both over time and geographically.

The package combines data-cleaning functions with custom
`ggplot2` timeline geoms and interactive `leaflet` maps.